---
title: "Analysis of Prenatal Care Visit Attendance and Preterm Birth"
author: "Eleane Lema"
date: "`r Sys.Date()`"
output: html_document
---

```{r, set-default-code-chunk-opts, echo = FALSE}
library(knitr)

opts_chunk$set(
  echo = FALSE,     # suppress code chunks
  warning = FALSE,  # suppress warnings
  message = FALSE  # suppress messages
)

```

# Introduction
This report analyzes a birth certificate dataset provided through the 
RSPH online course, Introduction to Data Analysis: A Self-Guided
Project. The dataset includes 80,015 births and has information on various 
maternal characteristics (race, ethnicity, age, 
smoking status), prenatal care use (month of first prenatal care visit, # of 
prenatal care visits), multiple births, gestational age at birth, and the 
version of the birth certificate.

The analysis objectives of this report are to:

1. Create a Table 1 to display birth characteristics stratified by prenatal
care visit attendance.
2. Plot the incidence of preterm birth (< 37 gestational age in weeks)
across varying levels of prenatal care visit attendance.
3. Fit a regression model to estimate the association between prenatal care
visit attendance and incidence of preterm birth after adjusting for smoking
status, maternal age, and race.


# Objective 1:
### Table 1: Birth characteristics

The sample included a total of 46,529 births with complete information on the
exposure, outcome, and covariates. Overall, 11% of births were preterm and
13% were to mothers who had smoked during their pregnancy. The majority of
births were to White mothers (82%) and mothers aged 20-29 years old (55%).

More than half (57%) of births were to mothers who attended less than 12
prenatal care visits. These mothers did not meet the recommended frequency of 12
to 14 visits according to [clinical guidelines](https://pmc.ncbi.nlm.nih.gov/articles/PMC10366963/).
Characteristics were similar between births with varying levels of prenatal care
attendance. However, mothers who attended less than 12 prenatal care visits
were more likely to experience preterm births compared to mothers who attended
more visits.


```{r, table-one}
here::i_am("report.Rmd")

table_one <- readRDS(
  file = here::here("output/table_one.rds")
)

table_one
```



# Objective 2:
### Bar plot of preterm birth incidence by prenatal care visit attendance

The bar plot below shows preterm birth incidence by the number of prenatal care 
visits attended. Mothers who met or exceeded the recommended number of visits
(at least 12) had preterm birth rates of about 7–8%. In contrast, mothers who
attended fewer than 12 visits had a preterm birth rate of approximately 14%,
nearly twice as high as those who met the recommendation.

```{r, bar-plot}
#| fig.align = "center",
#| out.width = "600px"

knitr::include_graphics(
  path = here::here("output/bar_plot.png")
)


```


# Objective 3:
### Table 2: Association between prenatal care visit attendance and preterm birth incidence

Below are the log binomial regression results for the effect of prenatal care
visit attendance on preterm birth incidence after adjusting for smoking status,
maternal age, and race. Compared to mothers who attended less than 12 prenatal
care visits, mothers who met or exceeded the recommended number of visits had
significantly lower risk of preterm birth. Mothers who completed 12 to 14 prenatal
care visits had 48% lower risk of preterm birth and mothers who completed 15 or
more visits had 37% lower risk of preterm birth.

```{r, regression-analysis}

regression_table <- readRDS(
  file = here::here("output/regression_table.rds")
)

regression_table

```

